% document
\documentclass[10pt,graphics,aspectratio=169,table]{beamer}
\usepackage{../code}
\usepackage{csquotes}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{pgfplots}
% theme
\usetikzlibrary{arrows}
\tikzstyle{line}=[draw] % here
\usetikzlibrary{decorations.pathmorphing}
\tikzset{arrow/.style={-latex, shorten >=.5ex, shorten <=.5ex}}

\usetheme{metropolis}
% packages
\title{Lesson 6}
\author{Christian Schwarz, Jakob Krebs}
\date{2.12.2019}
\begin{document}
\maketitle

\begin{frame}{Contents}
    \tableofcontents
\end{frame}


\section{File IO and Debugging}
\begin{frame}{Sources and Solutions}
    \begin{itemize}
        \item we publish all code written in this course at \url{https://github.com/jkrbs/c_lessons}
        \item we will publish example solutions of the tasks on same site
        \item send us questions or your solutions to c-lessons@deutschland.gmbh
    \end{itemize}
\end{frame}

\begin{frame}[fragile]{fopen and fclose}
    \code{stdio.h} provides the following functions to open and close a file:

    \begin{codeblock}
FILE* fopen (char* filename, char* mode);
int fclose (FILE* stream);

//example
FILE* test = fopen("test.txt", "w");
fclose(test);
    \end{codeblock}

    Filenames can either be absolute (\code{"/home/foo/bar.txt"}) 
    or relative (\code{"test.txt"}). 
    Relative paths are relative to the "current working directory".
    That is the current directory of your shell when you execute the program.
    Shells can usually change this directory using \code{cd} (change directory), 
    and display it using \code{pwd} (print working directory).


    \textbf{This is not necessarily the directory that the program executable lies in.}  
\end{frame}
    

\begin{frame}{file modes}
    The \code {"w"} mode in \code{fopen(filename, mode)} specifies that we
    only want to \textbf{w}rite to the file.
    There are multiple different modes available: 


    \begin{tabular}{|l|l|l|l|}
        \hline 
        \textbf{mode} & \textbf{access} & \textbf{if file exists} & \textbf{if file doesn't exist}\\ \hline 
        \textbf{r} & read-only & read from start & return NULL \\\hline
        \textbf{w} & write-only & overwrite contents & create new \\\hline
        \textbf{a} & write-only & append & create new \\\hline
        \textbf{r+} & read+write & read from start, overwrite & return NULL \\\hline
        \textbf{w+} & read+write & read from start, overwrite & create new \\\hline
        \textbf{a+} & read+write & read from start, but append at the end& create new \\\hline
    \end{tabular}
\end{frame}

\begin{frame}[fragile]{"File"}
    
    \begin{itemize}
        \item  \code{FILE*} can be thought of as a pointer to a FILE structure managed
        by the C standard library that remembers all the necessary information
        to interact with the file. 

        \item  "File" should not be taken too literally here.
        Stream might have been the better term.
        For example, \code{stdin} and \code{stdout} are also \code{FILE*}s.

        \item Streams or "\code{FILE}"s really just represent an object that bytes 
            can be written to and / or read from.
    \end{itemize}
   
    
  

  

\end{frame}

\begin{frame}[fragile]{fread and fwrite}
    \begin{small}

    \begin{codeblock}[numbers=none, basicstyle=\small]
size_t fread(void* buffer, size_t size, size_t count, FILE* stream);
size_t fwrite(void* buffer, size_t size, size_t count, FILE* stream);
    \end{codeblock}

    \begin{itemize}
        \item \code{fread} reads bytes from the \code{stream} and writes them into 
        \code{buffer}.  

        \item \code{fwrite} reads bytes from \code{buffer} and writes them out to the 
        \code{stream}. 
    
    \end{itemize}

    The functions read/write \code{size} bytes for up to \code{count} times, or until 
    the stream has no more contents.

    They return the number of elements 
    (of size \code{size}) successfully read/written.

    Sometimes this is useful, e.g. if we want to read up to 20 \code{int}s:
    \begin{codeblock}[numbers=none, basicstyle=\small]
size_t ints_read = fread(buffer, sizeof(int), 20, file);
    \end{codeblock}

    But mostly we use them like this:
    \begin{codeblock}[numbers=none, basicstyle=\small]
size_t bytes_read = fread(buffer, 1, sizeof(buffer), file);
    \end{codeblock}

    \end{small}
\end{frame}

\begin{frame}[fragile]{file io example}
    
    \begin{codeblock}
FILE* logfile = fopen("log.txt", "a+");
// very unlikely to fail since "a+" creates nonexistant files
assert(logfile != NULL);

char buffer[1024]; 
do{
    size_t size = fread(&buffer, 1, sizeof(buffer), logfile);
    display_log(&buffer, size); // use the data
} while(size > 0);

char* msg = "we accessed the log file\n";
size_t size = fwrite(msg, strlen(msg), 1, log);
assert(size == 1); // was our data written successfully ?

fclose(config);
    \end{codeblock}

\end{frame}

\section{Debugging}
\begin{frame}[fragile]{Debugging}
    There's multiple possibilities why a program doesn't work as intended.
    As we discussed, the broad classification is between.
    \begin{itemize}
		\item Compiletime (+ link time) errors
		\item Runtime errors ( also called \textit{bugs})
	\end{itemize}\ \\\ \\
	\textit{Compiletime errors} are easily handable since the compiler shows you where and what they are\\\ \\
    \textit{Bugs} are oftentimes much harder to find because they could be 
        anywhere in your program and nobody warns you.
\end{frame}

\begin{frame}{Different kinds of Bugs}
	Bugs can appear due to different reasons
	\begin{itemize}
		\item Variable overflow
		\item Division by zero
		\item Infinite loops / recursions
		\item Range excess
		\item Segmentation fault
		\item Dereferencing \code{NULL} (or other invalid) pointers
		\item ...
	\end{itemize}
\end{frame}

\begin{frame}[fragile]{The GNU Debugger (gdb) }
	There are tools helping with bugs, called debuggers. GDB is one of them.\\
	\bigskip
	To use it
	\begin{itemize}
		\item You have to install the package \textit{gdb}\\
		\item You have to compile your program with the \textit{-g} flag
		\begin{lstlisting}[numbers=none]
$ gcc -g main.c
\end{lstlisting}
		\item After that you can start your program with gdb:
		\begin{lstlisting}[numbers=none]
$ gdb a.out
\end{lstlisting}
	\end{itemize}
\end{frame}

\begin{frame}[fragile]{Using gdb}
    \begin{lstlisting}
$ gdb -g intermediate_06_asciidungeon.c
$ gdb a.out
(gdb) start
Temporary breakpoint 1, main (argc=1, argv=0x7fffffffe028) at ...
41          player = init_entity(5, 8, 100, 'J');
(gdb) next
42          monster1 = init_entity(2, 3, 100, '*');
(gdb) step
init_entity (x_pos=2, y_pos=3, health=100, symbol=42 '*') at ...
104         struct entity *new_ent = malloc(sizeof new_ent);
(gdb) backtrace 
#0  init_entity (x_pos=2, y_pos=3, health=100, symbol=42 '*') at ...
#1  0x0000555555554938 in main (argc=1, argv=0x7fffffffe028) at ...
(gdb)    
\end{lstlisting}

\end{frame}
\begin{frame}{Commands}
	\begin{itemize}
        \item If you started gdb without a file you can load it with
            \textbf{file} \textit{file\_name}.
		\item Use \textbf{r[un]} to execute the program with gdb.\\
            If you have a segfault, it's a good idea to begin with that.
            It will give you further information about the crash location.
        \item If you want to debug from the beginning use \textbf{sta[rt]}
            to run and immediately break\\
        \item You can set an arbitrary amount of breakpoints
            with \textbf{b[reak]} \textit{line\_number} or \textbf{b[reak]} \textit{function\_name}.\\
		    Begin with a breakpoint at the point right before program crashes.
		\item Print values with \textbf{p[rint]} \textit{identifier}.
        \item Use \textbf{w[atch]} \textit{identifier} to break and print
            a variable when it's changed.
	\end{itemize}
\end{frame}
\begin{frame}{Once you're at a breakpoint}
	\begin{itemize}
		\item Use \textbf{n[ext]} to execute the next program line only.
		\item \textbf{s[tep]} executes the next instruction.
		\item You can jump to the next breakpoint with \textbf{c[ontinue]}.
		\item To see how you have come to this point in the program flow, type \textbf{backtrace} or \textbf{bt}.\\
		This shows you all functions you called to come there.
		\\\ \\ \\
		\item By only hitting the \textit{return key}, you repeat the last entered  command.
	\end{itemize}
	
\end{frame}
\begin{frame}[fragile]{Conditional breakpoints}
After setting a breakpoint, GDB assigns an ID to it.\\
You can use this ID to extend the functionality of that breakpoint.
	\begin{itemize}
		\item \textbf{con[dition]} \textit{breakpoint\_ID expression} sets a condition for your Breakpoint:
		\begin{lstlisting}[numbers=none,language=bash]
(gdb) br 42
Breakpoint 1 at 0xbada55: file main.c, line 42.
(gdb) condition 1 i@=@@=@3
        \end{lstlisting}
		\item For string comparison, set the string before comparing with \textbf{strcmp}:
		\begin{lstlisting}
(gdb) br main.c:42
Breakpoint 13 at 0xdeadbeef: file main.c, line 42.
(gdb) set $string_to_compare = "lolwut"
(gdb) cond 13 strcmp ( $stringtocompare, c ) @=@@=@ 0
        \end{lstlisting}
        \item use \textbf{con[dition]} \textit{breakpoint\_ID} to remove the condition:
	\end{itemize}
\end{frame}

\begin{frame}[fragile]{gdb --tui Mode}
    If you want a nicer interface where you can see multiple lines of your program,
    use 
    \begin{codeblock}
gdb --tui a.out
    \end{codeblock}

\end{frame}

\begin{frame}{useful gdb commands}
    \begin{tabular}{|l|l|}
        \hline
        \textbf{file} & load program\\\hline
        \textbf{r[un]} & execute program\\\hline
        \textbf{b[reak]} & set breakpoint\\\hline
        \textbf{sta[rt]} & execute program and break immediately\\\hline
        \textbf{p[rint]} & print variable\\\hline
        \textbf{w[atch]} & break and print variable when it changes\\\hline
        \textbf{n[ext]} & execute next line and break\\\hline
        \textbf{s[tep]} & execute next instruction and break\\\hline
        \textbf{c[ontinue]} & execute until next breakpoint\\\hline
        \textbf{backtrace} / \textbf{bt} & How did I end up here?\\\hline
    \end{tabular}
\end{frame}

\begin{frame}[fragile]{Task: Bughunting in the Ascii Dungeon}
    We (or more precisely, the fsr) prepared a little ascii dungeon
    littered with bugs. You can find it at. 

    \small
    \url{https://jkrbs.github.io/c_lessons/tasks/intermediate_06_asciidungeon.c}
    
    Or just click on \code{Lesson 6 Intermediate Task: Bughunting in the Ascii Dungeon}
    on our website (\url{https://jkrbs.github.io/c_lessons}).


    Task: Fix all the bugs in the program using \code{gdb}, until you can
    run around on the Screen using wasd (+ Enter)! 
\end{frame}

\end{document}
